4.2: Difference-in-Difference Models

### 2019-12-05 static/slides/23-slides.Rmd
Update lesson 23 content

---
title: "4.2: Differences-in-Differences Models"
subtitle: "ECON 480 · Econometrics · Fall 2019"
author: 'Ryan Safner<br> Assistant Professor of Economics <br> <a href="mailto:safner@hood.edu"><i class="fa fa-paper-plane fa-fw"></i> safner@hood.edu</a> <br> <a href="https://github.com/ryansafner/metricsf19"><i class="fa fa-github fa-fw"></i> ryansafner/metricsf19</a><br> <a href="https://metricsF19.classes.ryansafner.com"> <i class="fa fa-globe fa-fw"></i> metricsF19.classes.ryansafner.com</a><br>'
#date:
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML" # rescales math with css changes https://github.com/yihui/xaringan/issues/143
    lib_dir: libs
    df_print: paged
    #seal: false
    css: [custom.css, custom-fonts.css, "hygge"] #, metropolis, metropolis-fonts
    nature:
      beforeInit: ["macros.js", "https://platform.twitter.com/widgets.js"] # first is for rescaling images , second is for embedding tweets, https://github.com/yihui/xaringan/issues/100
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
    includes:
      in_header: header.html # for font awesome, used in title  
---

@@

@@

- Often, we want to examine the consequences of a change, such as a law or policy
    - e.g. States that implemented law $X$ saw a change in $Y$
    - **Treatment**: States that implement law $X$
    - **Control**: States that did not implement law $X$
    - If we have **panel data** with observations for all states before *and* after the change:
- Simple logic: compare difference in outcomes of treatment group (before and after treatment) with those of non-treated group (before and after same treatment period)

@@

.center[
![](https://www.dropbox.com/s/vpk560fok07yo9t/dndhot.jpg?raw=1)
]

# Difference-in-Difference Models III

- The .shout[difference-in-difference model] (aka .shout["diff-in-diff"] or .shout[DND"]) identifies treatment effect by differencing the difference pre- and post-treatment between treatment and control groups

@@

class: inverse. center, middle

>>>>

---
title: "4.2: Difference-in-Difference Models"
subtitle: "ECON 480 · Econometrics · Fall 2019"
author: 'Ryan Safner<br> Assistant Professor of Economics <br> <a href="mailto:safner@hood.edu"><i class="fa fa-paper-plane fa-fw"></i> safner@hood.edu</a> <br> <a href="https://github.com/ryansafner/metricsf19"><i class="fa fa-github fa-fw"></i> ryansafner/metricsf19</a><br> <a href="https://metricsF19.classes.ryansafner.com"> <i class="fa fa-globe fa-fw"></i> metricsF19.classes.ryansafner.com</a><br>'
#date:
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML" # rescales math with css changes https://github.com/yihui/xaringan/issues/143
    lib_dir: libs
    df_print: paged
    #seal: false
    css: [custom.css, custom-fonts.css, "hygge"] #, metropolis, metropolis-fonts
    nature:
      beforeInit: ["macros.js", "https://platform.twitter.com/widgets.js"] # first is for rescaling images , second is for embedding tweets, https://github.com/yihui/xaringan/issues/100
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
    includes:
      in_header: header.html # for font awesome, used in title  
---

@@

# Clever Research Designs Identify Causality

```{r, fig.retina=3, fig.align="center", fig.width=13}
ggplot(data = tibble(x=c(0,10),
                     y=c(0,10)))+
  aes(x = x,
      y = y)+
  geom_label(x=1,y=2, label="Correlation", size =10, color = "#fde0dd")+
  geom_label(x=5,y=2, label="Causation", size = 10, color = "#7a0177")+

  geom_label(x=1,y=4, label="Differences", size =5, color = "#fde0dd")+
  geom_label(x=1,y=3.5, label="Pre-Post", size =5, color = "#fde0dd")+
  geom_label(x=2,y=4.5, label="Multiple Regression", size =5, color = "#fbb4b9")+
  geom_label(x=2,y=3.5, label="Matching", size =5, color = "#fbb4b9")+
    geom_label(x=3,y=5, label="Fixed Effects", size =5, color = "#f768a1")+
    geom_label(x=3.5,y=4.5, label="Diff-in-Diff", size =5, color = "#c51b8a")+
  geom_label(x=4,y=3.5, label="Natural Experiments", size =5, color = "#c51b8a")+
  geom_label(x=3.5,y=4, label="Regression Discontinuity", size =5, color = "#c51b8a")+
  geom_label(x=5,y=4.5, label="RCTs", size =5, color = "#7a0177")+
  annotate("segment", x = 1, xend = 5, y = 3, yend = 3, colour = "black", size=2, alpha=1, arrow=arrow(length=unit(0.5,"cm"), ends="last", type="closed"))+
  scale_x_continuous(breaks=seq(0,6,1),
                     limits=c(0,6))+
  scale_y_continuous(breaks=seq(1,5,1),
                     limits=c(1,5))+
  theme_void()

```

Again, this toolkit of research designs to identify causal effects is the economist's comparative advantage that firms and governments want!

@@

- Often, we want to examine the consequences of a change, such as a law or policy

- Often, we want to examine the consequences of a change, such as a law or policy

- .green[**Example**]: how do States that implement law $X$ see changes in $Y$
  - **Treatment**: States that implement law $X$
  - **Control**: States that did not implement law $X$

- Often, we want to examine the consequences of a change, such as a law or policy

- .green[**Example**]: how do States that implement law $X$ see changes in $Y$
  - **Treatment**: States that implement law $X$
  - **Control**: States that did not implement law $X$

- If we have **panel data** with observations for all states before *and* after the change...

- Find the *difference* between treatment & control groups *in* their *differences* before and after the treatment period

.pull-right[
.center[
![](https://www.dropbox.com/s/vpk560fok07yo9t/dndhot.jpg?raw=1)
]
]

@@

- The .shout[difference-in-difference model] (aka .shout["diff-in-diff"] or .shout[DND"]) identifies treatment effect by differencing the difference pre- and post-treatment values of $Y$ between treatment and control groups

@@
